Experience Report: Haskell in Computational Biology

### 2012-03-05 mrfy_experience_report.tex
tweak viterbi types, rearrange implementation section

\subsection{The algorithms}

@@

The resulting match states, as well as \textit{insertion} states that allow 
additional residues to appear in a protein sequence, and \textit{deletion} 
states that allow match states to be skipped, form a directed graph, the hidden 
Markov model (Figure~\ref{plan7}). Each match or insertion state has an 
associated table of emission probabilities, which correspond to the probability 
of seeing each specific amino acid letter at a given position. Triplets of 
insertion, deletion, and match states are grouped into ``nodes,'' which 
correspond to columns of the original alignment. What looks like an 
intimidating geometry is thus completely explained by the fact that we have 
insertions, deletions, and substitutions due to evolution.

@@

%% now explain computing a *parse* (HMMER) leading into Viterbi.

SMURF, SMURFLite, and MRFy all attempt to compute the optimal parse, or 
alignment, of a protein sequence onto the directed graph representing a Markov 
random field (see Figure \ref{mrf}). A parse is a mapping of the letters of a 
query sequence onto the states of a markov Random field (and the corresponding path through the finite state automaton), while an optimal parse 
is the parse exhibiting the greatest probability. The algorithm underlying 
MRFy grew out of an understanding of the limitations of the SMURF (Structural 
Motifs Using Random Fields) algorithm, which used multidimensional dynamic 
programming to compute the optimal parse of a query sequence onto a Markov 
random field. Rather than attempt to compute exactly the optimal parse, MRFy 
treats the non-local interactions (which correspond to hydrogen-bonded ``beta 
strands'' in the protein) as ``beads'' which can be slid along a string. In 
this analogy, the string is the sequence of amino acids of the query protein. A 
given placement of these beads corresponds to an assignment of particular amino 
acid residues being hydrogen bonded to other particular amino acid residues, 
and thus MRFy computes a probability based on observed frequencies of residues 
hydrogen bonding in beta strands~\cite{betawrappro}. In between these beta 
strands, the Markov random field is in fact a traditional hidden Markov model, 
whose optimal solution can be easily computed via dynamic programming (known as 
the Viterbi algorithm). MRFy performs a stochastic search to place these beta 
strands, based on an objective function that combines the probability computed 
by the Viterbi algorithm with that computed from the paired, hydrogen-bonded 
residues in the beta strands.

%% "nailing down" certain subsequences of the model; between beta strands we have HMMs

\begin{figure} 
\centerline{\includegraphics[width=8cm]{mrf_interleave_diagram.pdf}} 
\caption{The directed graph representing the states of a MRFy Markov random 
field finite state automaton. Shaded nodes represent beta-strand positions which have been fixed into match states, 
while the dashed edges represent the long-range dependencies due to hydrogen 
bonding. Note that in between the beta-strand regions are normal HMMs.}\label{mrf} \end{figure}

The Viterbi algorithm~\cite{Viterbi:1967} comprises a set of recurrence 
relations (see Equation \ref{viterbi_eqn}) that seek to maximize the 
probability of an observed sequence (in this case, of amino acid ``letters'') 
being emitted by a particular finite-state automaton (the hidden Markov model). 
The particular states of this finite-state automaton map to the recurrence 
relations, and in this instance derive from what is known in the computational 
biology community as a \textit{profile} hidden Markov model. The Viterbi 
algorithm in this instance recurses on ``insertion,'' ``deletion,'' and 
``match'' states (Figure~\ref{plan7}). Insertion states represent mutations 
that gain an amino acid letter in the query sequence, deletion states represent 
mutations that lose an amino acid letter in the query sequence, and match 
states represent either conserved positions (no change) or mutations that 
substitute one amino acid for another. The match and insertion states each have 
an associated table of emission probabilities, determined during training, 
indicating how likely that state is to emit a particular amino acid letter. 
Edges from state to state represent state transitions, and have associated 
probabilities determined during training. To be efficient, the Viterbi 
algorithm relies on dynamic programming, or memoization, and the resulting 
asymptotic complexity is $O(|M|\times|N|)$, when $M$ is the sequence of 
distinct states (nodes in the automaton's graph) and $N$ is the sequence of 
letters in the protein sequence. In our particular problem domain, both $M$ and 
$N$ routinely contain several hundred to a few thousand elements. The types of 
these states and probabilities are provided in Figure~\ref{viterbi_types}.

%% break the viterbi types up and insert into the Plan7 explanation.

\begin{figure*}
\lstinputlisting[label=viterbi_types,caption=Viterbi Types]{viterbi_types.hs}
\end{figure*}

%% group nodes, illustrate an alignment

@@

>>>>

\subsection{The Viterbi Algorithm}

@@

The resulting match states, as well as \textit{insertion} states that allow 
additional residues to appear in a protein sequence, and \textit{deletion} 
states that allow match states to be skipped, form a directed graph, the hidden 
Markov model (Figure~\ref{plan7}). Each match or insertion state has an 
associated table of emission probabilities, which correspond to the probability 
of seeing each specific amino acid letter at a given position. Each edge in the 
graph has a weight which is the probability of the associated state transition. 
Triplets of insertion, deletion, and match states are grouped into ``nodes,'' 
which correspond to columns of the original alignment. What looks like an 
intimidating geometry is thus completely explained by the fact that we have 
insertions, deletions, and substitutions due to evolution.

@@

If a biologist has trained a hidden Markov model on a group of related 
proteins, she may next use HMMER to determine the likelihood that a new 
protein, whose sequence is known but structure and function are not, shares 
structure and evolutionary history with the proteins that comprise the hidden 
Markov model. HMMER attempts to compute the optimal \textit{parse}, or 
alignment, of a protein sequence onto the directed graph of the hidden Markov 
model. A parse is a mapping of the letters of a query sequence onto the states 
of the model; equivalently, a parse is a \textit{path} through the edges of the 
graph. Since each edge has a weight which is a probability, and each emitting 
state (match or insertion) has a table of emission probabilities, the 
probability of a particular path through the model is simply the product of the 
associated probabilities. This path and its associated probability are computed 
using the Viterbi algorithm~\cite{Viterbi:1967}.

The Viterbi algorithm comprises a set of recurrence 
relations (see Equation \ref{viterbi_eqn}) that seek to maximize the 
probability of an observed sequence (in this case, of amino acid ``letters'') 
being emitted by a particular finite-state automaton (the hidden Markov model). 
To be efficient, the Viterbi 
algorithm relies on dynamic programming, or memoization, and the resulting 
asymptotic complexity is $O(|M|\times|N|)$, when $M$ is the sequence of 
distinct states (nodes in the automaton's graph) and $N$ is the sequence of 
letters in the protein sequence. In our particular problem domain, both $M$ and 
$N$ routinely contain several hundred to a few thousand elements. The types of 
these states and probabilities are provided in Figure~\ref{viterbi_types}.

%% break the viterbi types up and insert into the Plan7 explanation.
%% go over this with NR; I can't make it flow.

\begin{figure}
\lstinputlisting[label=viterbi_types,caption=Viterbi Types]{viterbi_types.hs}
\end{figure}

@@

\subsection{MRFy Implementation}

HMMER simply implements the Viterbi algorithm for computing the optimal parse 
of a protein sequence onto a hidden Markov model. Thus, the graphical model of 
a HMMER hidden Markov model is exactly that shown in Figure~\ref{plan7}. In 
contrast, MRFy builds on the approach of SMURF~\cite{smurf} and 
SMURFLite~\cite{smurflite}, our previous work. Like SMURF and SMURFLite, MRFy 
pays attention to longer-range structural interactions than a simple hidden 
Markov model can understand. In particular, these approaches ``nail down'' 
certain subsequences of the model, corresponding to the non-local hydrogen-bond 
interactions between beta strands. Beta strands are regions of protein 
structure that ``stick'' to one another. Because of their pairwise 
interactions, they are highly conserved over the course of evolution. Moreover, 
again because of these pairwise interactions, a change in one amino acid in a 
beta strand may require a change in a corresponding, hydrogen-bonded residue in 
another beta strand if the protein is to remain stable. The architecture of the 
resulting graphical model, derived from the HMMER ``Plan7'' architecture, is 
illustrated in Figure~\ref{mrf}.

These non-local interactions cause what had been a simple hidden Markov model 
to become a Markov random field. SMURF and SMURFLite attempted to compute the 
optimal parse of a protein sequence onto the directed graph representing a 
Markov random field using multidimensional dynamic programming. SMURF's 
computational complexity quickly exploded with complex protein structures, 
while SMURFLite limited its computational complexity by also limiting the 
complexity of protein structure it would understand. The algorithm underlying 
MRFy grew out of an understanding of these limitations.

Rather than attempt to compute exactly the optimal parse, MRFy treats the beta 
strands as ``beads'' which can be slid along a string. In this analogy, the 
string is the sequence of amino acids of the query protein. A given placement 
of these beads corresponds to an assignment of particular amino acid residues 
being hydrogen bonded to other particular amino acid residues, and thus MRFy 
computes a probability based on observed frequencies of residues hydrogen 
bonding in beta strands~\cite{betawrappro}. In between these beta strands, the 
Markov random field is in fact a traditional hidden Markov model, whose optimal 
solution can be easily computed via the Viterbi algorithm. MRFy performs a 
stochastic search to place these beta strands, based on an objective function 
that combines the probability computed by the Viterbi algorithm with that 
computed from the paired, hydrogen-bonded residues in the beta strands. If a 
structure has many beta strands, then it also has many miniature hidden Markov 
models which can be solved via the Viterbi algorithm.

\begin{figure}[h!] 
\centerline{\includegraphics[width=8cm]{mrf_interleave_diagram.pdf}} 
\caption{The directed graph representing the states of a MRFy Markov random 
field finite state automaton. Shaded nodes represent beta-strand positions which have been fixed into match states, 
while the dashed edges represent the long-range dependencies due to hydrogen 
bonding. Note that in between the beta-strand regions are normal HMMs.}\label{mrf} \end{figure}
